They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-21 paper/paper.tex
incorporate some andy feedback, start writing ww run streak stuff

The data set consists of all SnD matches played in tournanaments and
qualifiers during the CDL era, totaling 7,792 rounds across 852 series.
Data was collected in spreadsheets by community member
``IOUTurtle''.\footnote{Data: \url{https://linktr.ee/CDLArchive}.
  Author: \url{https://twitter.com/IOUTurtle}}

The empirical offensive round win percentage across all rounds is
47.8\%.\footnote{Offensive round win percentage has been nearly constant
  across the three games during the CDL era: 1. 47.2\% in MW (2020) 2.
  47.9\% in Cold War (2021) 3. 48.1\% in Vanguard (2022)} Table
\ref{tbl:cod-o-win-prop-by-series-state} shows round win percentages by
series ``state'' (i.e.~the number of round wins by each team prior to an
upcoming round). Offensive round win rate is not quite constant,
although never veers more than 10\% from this global average.

@@

Additionally, both Counter-Strike and Valorant have overtime rules and
blocked offensive/defensive roles (i.e.~playing either offense or
defense for many consecutive rounds). On the other hand, teams in CoD
SnD rotate sides every round, analogous to a 1-1-1-1-1-1-1 format for
home advantage in best-of-seven series for professional sports like the
MLB, NBA, and NHL.\footnote{1-1-1-1-1-1-1 is not used today in these
  leagues, but it was at least one in each league.} While theoretically
one might be able to account for any kind of balanced rotation, such as
a ``blocky'' on like a 5-5-1, the rotation of team sides every round is
convenient, particularly for convincing ourselves that rounds could
reasonably be modeled as i.i.d. Bernoulli trials.

@@

\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(r - s)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
}\label{eq:series-length}\end{equation}

\(s = 11\) rounds for CoD SnD.\footnote{\(\phi\) and other symbols are
  selected so as to avoid reserve symbols, like \(p\) for usage in other
  contexts, without risking confusion. Upper-case symbol, e.g.~\(\Phi\),
  are consistently used in this paper to represent proportions, while
  lower-case symbols corresponding to proportions represent
  probabilities. ``Hats'', e.g.~\(\hat{\Phi}\), are used to convey
  expectations, while ``bare'' symbols mirroring hat-ed symbols convey
  observational data.}

@@

\begin{equation}\protect\hypertarget{eq:pwr}{}{
p_0(\text{win} | r) = p^{w|r}_0 = \{
\begin{array}{rcll}
p^{w|r_W} & = & \frac{m}{r}, & \text{team wins series} \\
p^{w|r_L} & = & \frac{r - m}{r}, & \text{team loses series}
\end{array}
}
\label{eq:pwr}
\end{equation}

using \(m\) from Equation \ref{eq:m}. The Bernoulli round loss
probability for a series lasting \(r\) rounds, \(p^{\ell|kr}_0\)
(i.e.~\(\ell = 1 - w, w \in [0, 1]\)), can be formulated symmetrically.

As shown by MS, we should expect the proportion of rounds wins
immediately following a streak of \(k\) rounds wins for a series lasting
\(r\) rounds, \(\hat{P}^{w|kr}_{MS}\), to be strictly less than
\(p^{w|r}\).\footnote{See Appendix E.1 of MS for details on the proof
  that follows.}\footnote{Here, I choose symbols, e.g.~\(\pi\), \(x\),
  etc. that mostly do not conflict with syntax in other places. The
  exceptions are \(k\) and \(r\), which are used to represent streak
  length and number of trials globally.}

\begin{quote}
Theorem 1: \emph{Let}
\(\mathbf{X} = \{X_i\}^{r}_{i=1}, r \geq 3\)\emph{, be a sequence of
independent Bernoulli trials, each with probability of success}
\(0 < \pi < 1\)\emph{. Let} \(\hat{\Pi}_k(\mathbf{X})\) \emph{be the
proportion of successes on the subset of trials} \(I_k(\mathbf{X})\)
\emph{that immediately follow} \(k\) consecutive successes, that is,
\(\hat{\Pi}_k(\mathbf{X}) := \sum_{i \in I_k(\mathbf{X})} X_i | I_k(\mathbf{X}) |. \hat{\Pi}_k\)
\emph{is a biased estimate of}
\(\mathbb{P}(X_t = 1 | \prod_{j=t-k}^{t-1} X_j = 1) \equiv \pi\)
\emph{for all} \(k\) \emph{such that} \(1 \leq k \leq r - 2\)\emph{. In
particular,}
\(E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset] < \pi.\)
\end{quote}

MS notes that there does not exists a closed form representation for
\(E[\hat{\Pi}_k(\mathbf{X}) | I_k(\mathbf{X}) \neq \emptyset]\),
henceforth \(E^\Pi\), for \(k \> 1\). One may recursively build a
collection of dictionaries to quantify \(E^\Pi\) analytically, or, more
reasonably for large values of \(n\), one may run simulations to
estimate the expected value. Indeed, Figure 1 uses simulation to
estimate \(E^\Pi\) as a function of the total number of trials \(r\),
given \(\pi\) and \(k\).\footnote{I've adapted the code from Vafa
  (2017), which implements MS's framework (2018).}\footnote{Note that
  \(\pi\) and \(\hat{\Pi}_k(\mathbf{X})\) are analogues for
  \(p_0^{w|r}\) and \(\hat{P}^{w|kr}_{MS}\) in my notation for CoD SnD.}

\begin{figure}
\centering
\includegraphics{images/pwkr_ms.png}
\caption{The expected value of the proportion of successes on trials, $E^\Pi$, that immediately follow $k$ consecutive successes for Bernoulli trial success $\pi = 0.5$ and streak lengths $k \in [2, 3, 4]$, as a function of the total number of trials $r$.}
\end{figure}

My context is fundamentally different from that of GVT and MS, both of
whom focus on longitudinal data in controlled settings. GVT also
performs statistical tests on shots from players in live games,
i.e.~``observational'' data, but they note that their findings are
likely affected player shot selection in the face of defensive strategy
by the opposing team. The number of trials is fixed in their
experimental designs, but in CoD SnD, the number of rounds played is
determined as a function of the max number of possible rounds (\(s\))
and whether or not the team wins the series, as shown in Equation
\ref{eq:pwr}. The Bernoulli trial success probability, i.e.~a round win
in CoD SnD, is not independent of the opponent. Consequently, a
statistical test of the difference in \(\hat{P}^{w|kr}\) and
\(\hat{P}^{\ell|kr}\), as performed by GVT and MS to evaluate their
respective hypothesis regarding post-streak success rates is not
appropriate.

@@

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0(\text{win} | k, r) = p^{w|kr}_0 =\{
\begin{array}{rcll}
p_0^{w|kr_W} & = & \frac{m - k}{i - k}, & \text{team wins series} \\
p_0^{w|kr_L} & = & \frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}
}
\label{eq:pwkr}
\end{equation}

@@

If there are \(\sigma\) observed successes in a sample of \(r\) trials
and we expect that there should be \(r * \omega_0\), the probability of
getting this expected number of successes is

\begin{equation}\protect\hypertarget{eq:binom}{}{
\Pr(\sigma) = {\binom {r}{\sigma}} \omega^{\sigma}(1-\omega)^{r-\sigma}.
}\label{eq:binom}\end{equation}

Plugging in the notional probability \(p^{w|kr}_0\) for the null
\(\omega_0\) and the observed probability
\(p^{w|kr} = N^{w|kr} \times P^{w|kr}\), where \(N^{w|kr}\) is the
observed number of instances where a teams wins the round immediately
following a \(k\)-round winning streak in a series lasting \(r\) rounds,
for the observed \(\omega\), we can evaluate whether the observed
probability is not significantly different from the notional
probability. If we reject this hypothesis, then we might consider team
momentum, represented by post-streak success, plausible.

Returning to the streak-selection adjustment framework proposed by MS,
one can perform the same binomial test, swapping out \(p^{w|kr}_0\) with
\(p^{w|kr}_{MS}\). Of course, given the caveats mentioned before, the
results should be heeded with caution.

@@

>>>>

The data set consists of all SnD matches played in major tournaments and
qualifiers during the CDL era, totaling 7,792 rounds across 852
series.\footnote{288 of the series occur in major tournaments, which are
  considered to be more ``competitive'' than the qualifiers since they
  are played in person (COVID-permitting), whereas the qualifiers are
  played online.} Data was collected in spreadsheets by community member
``IOUTurtle''.\footnote{Data: \url{https://linktr.ee/CDLArchive}.
  Author: \url{https://twitter.com/IOUTurtle}}

The empirical offensive round win percentage across all rounds is
\(\phi_O\) = 47.8\%.\footnote{Offensive round win percentage has been
  nearly constant across the three games during the CDL era: 1. 47.2\%
  in MW (2020) 2. 47.9\% in Cold War (2021) 3. 48.1\% in Vanguard (2022)}
Table \ref{tbl:cod-o-win-prop-by-series-state} shows round win
percentages by series ``state'' (i.e.~the number of round wins by each
team prior to an upcoming round). Offensive round win rate is not quite
constant, although never veers more than 10\% from this global average.

@@

Additionally, both Counter-Strike and Valorant have overtime rules and
blocked offensive/defensive roles (i.e.~playing either offense or
defense for many consecutive rounds). On the other hand, teams in CoD
SnD rotate sides every round, analogous to a 1-1-1-1-1-1-1 format for
home advantage in best-of-seven series for professional sports like the
MLB, NBA, and NHL.\footnote{1-1-1-1-1-1-1 is not used today in these
  leagues, but it was at least one in each league.} While theoretically
one might be able to account for any kind of format, such as a 5-5-1,
the rotation of team sides every round is convenient for convincing
ourselves that rounds could reasonably be modeled as i.i.d. Bernoulli
trials.

@@

\begin{equation}\protect\hypertarget{eq:series-length}{}{
\hat{\Phi}(r) = \frac{(r - 1)!}{(m - 1)!(s - r)!}(\phi^{m}(1 - \phi)^{r - m} + \phi^{r - m}(1 - \phi)^m)
}\label{eq:series-length}\end{equation}

\(s = 11\) rounds for CoD SnD.\footnote{\(\phi\) and other symbols are
  selected so as to reserve symbols like \(p\) for usage in other
  contexts without causing confusion for the reader. Upper-case symbol,
  e.g.~\(\Phi\), are consistently used in this paper to represent
  proportions, while lower-case symbols corresponding to proportions
  represent probabilities. ``Hats'', e.g.~\(\hat{\Phi}\), are used to
  convey expectations, while ``bare'' symbols mirroring hat-ed symbols
  convey observational data.}

@@

\begin{equation}\protect\hypertarget{eq:pwr}{}{
p_0(\text{win} | r) = p^{w|r}_0 = \{
\begin{array}{rl}
\frac{m}{r}, & \text{team wins series} \\
\frac{r - m}{r}, & \text{team loses series}
\end{array}
}
\label{eq:pwr}
\end{equation}

using \(m\) from Equation \ref{eq:m}.

As shown by MS with their Theorem 1, we should expect the proportion of
rounds wins immediately following a streak of \(k\) rounds wins for a
series lasting \(r\) rounds, \(\hat{P}^{w|kr}_{MS}\), to be strictly
less than \(p^{w|r}\).\footnote{See Appendix E.1 of MS for details on
  the proof that follows.} Unfortunately, there does not exist a closed
form representation for the expectation for \(\hat{P}^{w|kr}_{MS}\) for
\(k > 1\). Nonetheless, one may recursively build a collection of
dictionaries to compute the expectation analytically, or, more
conveniently, one may run simulations to estimate the expected
value.\footnote{I've adapted the code from Vafa (2017), which implements
  MS's framework (2018).}

Although it would be convenient to emulate the hypothesis testing for
difference of proportions performed by MS (and GVT), my context is
fundamentally different from that of MS, who focus on longitudinal data
in controlled settings.\footnote{GVT also performs statistical tests on
  shots from players in live games, i.e.~``observational'' data, but
  they note that their findings are likely affected player shot
  selection in the face of defensive strategy by the opposing team.} The
number of trials is fixed in their experimental designs, but in CoD SnD,
the number of rounds played is determined as a function of the max
number of possible rounds (\(s\)) and whether or not the team wins the
series, as shown in Equation \ref{eq:pwr}. The Bernoulli trial success
probability, i.e.~a round win in CoD SnD, is not independent of the
opponent. Consequently, a statistical test of the difference in
\(\hat{P}^{w|kr}\) and \(\hat{P}^{\ell|kr}\), as performed by MS to
evaluate their hypothesis regarding post-streak success rates is not
completely appropriate, although useful as a reference.

@@

\begin{equation}\protect\hypertarget{eq:pwkr}{}{
p_0(\text{win} | k, r) = p^{w|kr}_0 =\{
\begin{array}{rl}
\frac{m - k}{i - k}, & \text{team wins series} \\
\frac{s - m - k}{i - k}, & \text{team loses series}
\end{array}
}
\label{eq:pwkr}
\end{equation}

@@

If there are \(r_+\) observed successes in a sample of \(r\) trials and
we expect that there should be \(r * \omega_0\), the probability of
arriving at this expected number of successes is

\begin{equation}\protect\hypertarget{eq:binom}{}{
\Pr(r_+) = {\binom {r}{r_+}} \omega^{r_+}(1-\omega)^{r-\sigma}.
}\label{eq:binom}\end{equation}

Plugging in the notional probability \(p^{w|kr}_0\) for the null
\(\omega_0\) and the observed probability
\(p^{w|kr} = N^{w|kr} \times P^{w|kr}\), for the observed \(\omega\),
where \(N^{w|kr}\) is the observed number of instances where a teams
wins the round immediately following a \(k\)-round winning streak in a
series lasting \(r\) rounds, we can evaluate the null hypothesis that
the observed probability is equal to the notional probability. If we
reject this null hypothesis, then we might consider team momentum,
represented by post-streak success, plausible.

One can perform the same binomial test for the MS probability adjusting
for streak-selection bias, \(p^{w|kr}_{MS}\). Given the caveats
mentioned before, the results should be heeded with caution.

Stepping back from the one-sided perspective of a single team's
probability of winning and their streakiness, I can attempt to detect
the ``hot hand'' more broadly with the Wald-Wolfowitz runs test, a
common tool used in the field of quality control. Under the null
hypothesis, the number of runs in a sequence of \(r\) trials is a random
variable with the following mean \(\mu\) and variance \(\sigma^2\).

\begin{equation}\protect\hypertarget{eq:ww}{}{
\mu = \frac{2r_+r_-}{r} + 1, \sigma^2 = \frac{(\mu-1)(\mu-2)}{r-1}
}\label{eq:ww}\end{equation}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

For the sake of brevity, I did not delve into win rates and streaks
split by offensive or defensive role. For example, it would be
interesting to look at whether the round win rate is higher after a
streaks of three round wins when the team starts the streak winning as
the offensive team. This would mean that, in the round immediately
following the streak, the streaking team would be playing defense, where
teams are slightly more likely to win on average. On the other hand,
streaks of three starting with an offensive round win---theoretically,
given \(\phi_O\) = 47.7\% the expected frequency is
\(\phi_O^2 \times (1 - \phi_O) \approx 0.119\)---are likely to occur
less frequently than such streaks starting with a defensive
win---\(\phi_O \times (1 - \phi_O)^2 \approx 0.130\)---so I would want
to account for sample size differences.

Regarding future work, I've reached out to the Twitter user
``R11stats'', who privately tracks in-round player
engagements.\footnote{\url{https://twitter.com/R11stats}} R11stats
expressed intent on making the data public, which would allow for
research into player-specific momentum.

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

I would like to thank my brother Andrew for giving me feedback on early
iterations of this paper, and for requiring me to mention him.
